Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-07 survey/survey.tex
Make the use usage of sum constructor consistent.

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\alt}{~|~}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$
versus $\top$; the first has exactly \emph{two} proofs of inhabitation
($\inl{\texttt{tt}}$ and $\inr{\texttt{tt}}$) while the second only one
($\texttt{tt}$). These cannot be put in bijective correspondence. Even
though the ``payload'' \texttt{tt} is the same, forgetting
$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
we have expressly disallowed.  Yes, this should remind you of
Maxwell's daemon: even though the data is the same, they are tagged
differently, and these tags are indeed information, and their
information content must be preserved.

@@

\noindent
Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, choice
composition $\oplus$ (sometimes called juxtaposition), and parallel
composition $\otimes$. See Fig.~\ref{pi-combinators} for the
signatures. The construction $c_1 \odot c_2$ corresponds to performing
$c_1$ first, then $c_2$, and is the usual notion of composition -- and
corresponds to $\fatsemi$ of the language of permutations of
Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
construction $c_1 \otimes c_2$ operates on a product structure, and
applies $c_1$ to the first component and $c_2$ to the second. The
language of permutations lacked the ability to combine permutations by
taking sums and products, which led to the awkward non-compositional
programming style illustrated in the full adder
example~(Eq.~\ref{eq:adder}).

>>>>

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inleft}[1]{\textsf{left}~#1}
\newcommand{\inright}[1]{\textsf{right}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\alt}{~|~}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$ versus
$\top$; the first has exactly \emph{two} proofs of inhabitation
($\inleft{\texttt{tt}}$ and $\inright{\texttt{tt}}$) while the second
only one ($\texttt{tt}$). These cannot be put in bijective
correspondence. Even though the ``payload'' \texttt{tt} is the same,
forgetting $\texttt{left}$ (or \texttt{right}) throws away information
-- something we have expressly disallowed.  Yes, this should remind
you of Maxwell's daemon: even though the data is the same, they are
tagged differently, and these tags are indeed information, and their
information content must be preserved.

@@

\noindent
Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, choice
composition $\oplus$ (sometimes called juxtaposition), and parallel
composition $\otimes$. See Fig.~\ref{pi-combinators} for the
signatures. The construction $c_1 \odot c_2$ corresponds to performing
$c_1$ first, then $c_2$, and is the usual notion of composition -- and
corresponds to $\fatsemi$ of the language of permutations of
Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{left}$ or $\textsf{right}$ respectively. Finally the
construction $c_1 \otimes c_2$ operates on a product structure, and
applies $c_1$ to the first component and $c_2$ to the second. The
language of permutations lacked the ability to combine permutations by
taking sums and products, which led to the awkward non-compositional
programming style illustrated in the full adder
example~(Eq.~\ref{eq:adder}).
